From Stone to Tool: How Raw Materials Influenced Upper Paleolithic Technology in Southwestern Iberia (Vale Boi)

### 2025-04-15 analysis/paper/paper.qmd
Final commit before submission 03. Project is up to date with final documents.

---
title: "From Stone to Tool: How Raw Materials Influenced Upper Paleolithic Technology in Southwestern Iberia (Vale Boi)"
author:
  - Joana Belmiro:
      correspondence: "yes"
      email: jfbelmiro@ualg.pt
      orcid: https://orcid.org/0000-0003-4951-2907
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
  - Nuno Bicho:
      orcid: https://orcid.org/0000-0001-9655-0549
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
  - Xavier Terradas:
      orcid: https://orcid.org/0000-0002-8000-5607
      institute:
        - Institución Milá y Fontanals de Investigación en Humanidades, Consejo Superior de Investigaciones Científicas, Barcelona, Spain
  - João Cascalheira:
      orcid: https://orcid.org/0000-0003-0321-8892
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
title-block-published: "Last updated"
format:
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
number-sections: true
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/archaeological-and-anthropological-sciences.csl" # Insert path for the bib-style
abstract: |
  The Upper Palaeolithic of Westernmost Europe was marked by technological and cultural transformations shaped by Late Pleistocene climatic shifts. The Iberian Peninsula, particularly southwestern Iberia, served as a key eco-cultural niche, sustaining hunter-gatherer populations through environmental fluctuations. Vale Boi, the only site in southwestern Iberia with a nearly complete UP sequence, provides critical insights into technological, economical and mobility patterns over time. This study examines lithic raw material procurement and technological organization at Vale Boi, focusing on the role of non-local cherts in mobility strategies and social interactions. Using a combined techno-typological and geoarchaeological approach, we analyzed chert assemblages from Gravettian, Proto-Solutrean, and Solutrean levels, integrating previously published macroscopic and petrographic data. Our results indicate that non-local cherts were intensively reduced, suggesting strategic curation. The presence of diverse tool types challenges the assumption that non-local materials were transported solely as finished tools, instead highlighting their role as versatile raw material sources. Gravettian occupations exhibited greater reliance on non-local cherts, implying frequent mobility, while later phases saw increased dependence on local materials, aligning with more long-term occupations.These findings refine our understanding of UP technological adaptations in southwestern Iberia, emphasizing the evolving role of lithic resources in mobility, settlement, and social networks. By integrating lithic technology and raw material sourcing, this study contributes to broader discussions on eco-cultural niches and UP hunter-gatherer adaptations.

@@

In westernmost Europe, five main technocomplexes have been identified within the UP - Aurignacian, Gravettian, Proto-Solutrean, Solutrean, and Magdalenian - each contributing new evidence on resilient behaviors and cultural continuity. These technocomplexes are distinguished by regional adaptations as well as technological and subsistence innovations [@baker2024; @bicho2017; @bicho2011; @bicho2008; @cascalheira2017; @straus1991; @straus1993; @zilhão1997].

These innovations are believed to have resulted from human adaptations and resilience to the shifting conditions of the Late Pleistocene, which was marked by abrupt climatic events such as the Heinrich events and the Last Glacial Maximum (LGM) [@sanchezgoñi2010]. These climatic shifts had a profound impact on flora and fauna, with exception of spatially limited refugia, likely disrupting settlement patterns and the organization of hunter-gatherer groups, ultimately leading to a process of cultural reorganization [@bradtmöller2012; @morin2008; @schmidt2012], which has frequently been a focus of UP research.

@@

In this context, comparisons between Vale Boi and the Gravettian occupations in central Portugal and southern Spain, revealed marked regional differences, suggesting more limited movement of people and information in southwestern Iberia. These differences were suggested to be related to Vale Boi’s peripheral location and the distinct ecological characteristics of westernmost Algarve [@marreiros2013]. In contrast, Proto-Solutrean evidence from the site indicates technological similarities and strong connections with central Portugal [@belmiro2021], while the Solutrean period stands out as a highly dynamic phase at Vale Boi, during which the site functioned as a key connection point between central Portugal and southern Spain, facilitating social networks and cultural exchanges [@bicho2017a; @cascalheira2013c].

@@

To address this gap, this paper presents a combined analysis of lithic technology and raw material sourcing, aiming to refine our understanding of technological and social adaptations throughout the UP in southwestern Iberia. By integrating archaeological and ethnographic perspectives, we test expectations regarding the organisation of hunter-gatherer technological systems and their implications for mobility, resource management, and social interactions. In doing so, we contribute to ongoing discussions about eco-cultural niches and the role of southwestern Iberia within broader UP social networks.

@@

Lithic technological organization has long been a central focus in archaeology, offering critical insights into past human behaviour and adaptive strategies. It encompasses all stages of stone tool production and use, from raw material acquisition to discard. Nelson [-@nelson1991] defines it as the selection and integration of strategies for tool production, use, transport, and maintenance, while Kelly [-@kelly1988] emphasizes its spatial and temporal dimensions within cultural systems. These perspectives highlight the interrelationship between raw material sourcing, mobility, and settlement organization—key factors in understanding hunter-gatherer adaptations.

@@

Vale Boi is located on the western coast of the Algarve region, southern Portugal, at a latitude of N 37º5'24.1" and longitude of W 8º48'32.4" (@gibajaFUNCIONINSTRUMENTOSLITICOS2006; the original site coordinates were converted from Degree/Minute to Degree/Minute/Second), within a small valley extending \~2 km southward to the Atlantic Ocean [@fig-map a]. The site is nestled among limestone outcrops that create west- and southwest-facing rock shelters [@fig-map b]. Covering more than 10,000 square meters along the valley slope, Vale Boi includes three main excavated areas, Slope, Terrace, and Shelter [@fig-map c], which were excavated between 2000 and 2019 [@bicho2007; @bicho2012; @bicho2013; @cascalheira2017; @cascalheira2013; @manne2012]. Based on the available raw material and technological evidence, particularly the macroscopic and petrographic data [@belmiro2025], the Terrace and Shelter areas were selected for the current study.

@@

The site’s lithic assemblages represent all stages of stone tool production, with raw material use varying by occupation. Technological analyses of these assemblages have provided insights into the adaptive strategies of UP hunter-gatherers at Vale Boi, revealing technological shifts across different cultural phases. Given the chosen sample for this study (see Section 2.2. Datasets and analysis), we summarise here the results for the technological analysis of the Solutrean assemblages (levels A-C) of the Shelter [@cascalheira2010; @cascalheira2013a] and the Proto-Solutrean assemblages (levels 4E-5) of the Terrace [@belmiro2020; @belmiro2021]. The Gravettian occupations (levels 6-8) of the Terrace have been recently studied, but the results remain unpublished.

@@

Previous studies showed that the main raw materials used at the site are chert, quartz, greywacke and smaller amounts of chalcedony, with the percentages of these changing across occupations. Notably, the lower Proto-Solutrean levels show high frequencies of quartz [@belmiro2021], whereas chert becomes the dominant raw material in the Solutrean occupations of the Shelter [@cascalheira2010]. Likewise, different raw materials appear to have been used for specific purposes,with chert being particularly associated with the production of high percentage of blanks and retouched tools, and greywacke being extensively used for anvils [@horta2019].

Quartz and greywacke are locally available from streambeds near the site, whereas regional chert and chalcedony sources lie further afield, within a \~20 km radius. Our previous studies showed, as expected, the existence of local cherts in the assemblages [@fig-cherts a], originating from the Lower Jurassic (cherts) and Triassic (chalcedony) outcrops in proximity to the site [@belmiro2025]. Despite petrographic differences between cherts and chalcedony, for this study any chert varieties (e.g., chalcedony or jasper) are referred to as chert [@luedtke1978]. A significant portion of non-local cherts was also identified [@fig-cherts b-c], with some identified sources, such as the Cretaceous outcrops of central Portugal (Lisbon and Rio Maior areas) and the Upper Jurassic oolitic/peloidal cherts of the Betic Systems in southern Spain. This indicates varying degrees of contact or long-range mobility throughout the UP, likely influenced by shifts in settlement patterns and the duration of site occupation.

@@

Cascalheira et al. (2017) applied the Whole Assemblage Behavioral Index (WABI) [@riel-salvatore2004; @riel-salvatore2010] to reconstruct and compare mobility strategies at Vale Boi. By analysing lithic volumetric density and retouched tool frequencies from the Shelter’s Solutrean [@cascalheira2010; @cascalheira2013a] and Magdalenian [@mendonça2009] levels, as well as the Terrace’s Gravettian levels [@marreiros2015], the authors confirmed that most occupations corresponded to long-term residential base camps, composed of high-density assemblages with low degree of retouch, except for the Early Gravettian and Magdalenian occupations. Regarding the Terrace area, however, the results were based on previous excavations (\<2012), and the identification of new layers and reorganization of the occupations is yet to be tested through the use of lithic techno-typological methods.

@@

One challenge of integrating datasets collected by different researchers is variability in methodologies and recorded variables. However, this study minimizes such inconsistencies by ensuring methodological alignment across datasets, as all data collection followed standardized attribute-based lithic analysis protocols. The lithic analysis datasets used in this study adopted an attribute-based methodology, which aims to describe morphological and metrical attributes of technological classes. The attributes analysed followed those present in specialized literature, such as Brézillon [-@brézillon1968], and [@tixier1980], paired with other Palaeolithic lithic attribute analyses [e.g., @bicho1992; @zilhão1997; @cascalheira2019; @tostevin2012].

After data collection, the databases were imported into R Programming Environment, where the information was processed through the creation of descriptive statistical analysis and writing of this article. Some differences in the collection process, especially between the study by Cascalheira [-@cascalheira2010] and the following studies, required homogenization. For example, where in Cascalheira (2010) retouch was an independent variable (e.g., Class as flake, Retouch as Yes), in Belmiro et al. [-@belmiro2021], retouch was identified as Class variable. In this situation, the Solutrean dataset was modified, so that all artifacts with retouch were identified in the Class variable, instead of the original classification. All changes applied to the original data made for homogenisation between datasets can be consulted in the R Studio scripts in our online repository (<http://www.doi.org/10.17605/OSF.IO/UWMPK>).

To integrate technological and raw material data, we merged multiple datasets using the artifact identifier (“ID”), which corresponds to total station plot records. This identifier links artifacts to either precise excavation coordinates (for individually plotted pieces) or generalized spatial data (for sieved assemblages).

Given that the raw material analysis focused exclusively on individually piece-plotted artifacts (dimensions \>2 cm), the final dataset does not include all excavated material but retains the most diagnostically relevant pieces for analysis. As a result, the final dataset used in this study does not directly correspond to the full datasets analysed in the original studies [@tbl-samples]. Similarly, the raw material study used only currently accessible artifacts from the assemblages. This is especially relevant for the Shelter Solutrean chert assemblage, where shatter is absent from the sample due to the lack of availability for study.

@@

Following data homogenization and integration, we conducted descriptive statistical analyses and non-parametric statistical analyses to examine correlations between raw material use and technological patterns and test expectations related to the technological organisation of local and non-local cherts [see @tbl-expectations], and between different UP occupations. To test these expectations, we evaluated 9 parameters for each dataset and assemblage; the parameters and their calculation/measurement methods are detailed in @tbl-parameters.

@@

The entirety of the R code used for the analysis, datasets, and visual representations contained in this paper can be accessed through our online repository (<http://www.doi.org/10.17605/OSF.IO/UWMPK>). We used the rrtools package by Marwick et al. [-@marwick2018] to create a research compendium and write a reproducible journal article. The provided files include the complete set of raw data used in the analysis, along with a custom R project [@wickham2015] containing the code required to generate all tables and figures. To enable maximum reuse, the code is made available under the MIT license, data under CC‐0, and figures under CC‐BY (additional details can be found in @Marwick2017).

@@

Results are structured according to the predefined analytical parameters, comparing local and non-local cherts [@fig-cherts] across different assemblages. A comprehensive description of each assemblage and detailed tables for all analyses are provided in Online Resource 1.

Blanks and blank fragments represent the most abundant lithic class in all assemblages, accounting for over 40% of artifacts across all chert types [@fig-general-graphs]. As expected, cores are present in lower frequencies (\~3-8%) but are more numerous in local cherts than in non-local types and absent from the TL group. In the Proto-Solutrean assemblage, non-local chert cores are rare, with only T7-9 yielding a small number (n=4 complete cores, n=2 core fragments). Albeit in very small numbers in all assemblages (n\<10), core preparation and maintenance products are also present in local and non-local cherts, especially in the non-local types which show the larger sample number such as T6-7. The presence of knapping by-products such as shatter is also seen in both local and non-local cherts in the Gravettian [@fig-general-graphs a] and Proto-Solutrean [@fig-general-graphs b] assemblages. Retouched tools are present in all chert types across all assemblages. In general, local cherts exhibit lower retouch frequencies than non-local cherts, though the total number of retouched tools varies significantly between assemblages.

@@

The Gravettian assemblage exhibits the highest proportion of retouched tools [\~21%; @fig-general-graphs a], compared to \<10% in other assemblages. Tool diversity is greater among non-local cherts, which contain a high frequency of generalized tools (e.g., notches, retouched pieces) alongside hunting implements such as backed bladelets. The abundance of burins and burin spalls suggests they may have functioned as cores, indicating a technological strategy rather than solely a focus on retouch (Online Resource1 Table s2). In comparison to the Gravettian, other assemblages show smaller numbers of retouched tools. While in the Proto-Solutrean (Online Resource 1 Table s4) no significant difference was identified between local and non-local retouched tools, in the Solutrean assemblage (Online Resource 1 Table s6) both tool diversity and tool specialization are higher in local cherts.

Retouched tools were further explored through the tools-to-debitage ratio [@fig-ratios a, Online Resource 1 Table s8]. In general, the results show that non-local chert types have higher retouch-to-debitage ratios than local types, albeit with differences between assemblages. The Gravettian assemblage shows the highest tool-to-debitage ratios compared to the remaining assemblages; non-local cherts show ratios above 0.9, while local cherts show values below 0.5. In the Proto-Solutrean non-local cherts show ratios between 0.3 to 0.5, while local cherts show ratios below 0.3. This pattern is less clear in the Solutrean assemblage since the ratios are the lowest of all assemblages. Non-local cherts show ratios above 0.1 (with exception of T8) while local cherts show ratios below 0.1 (with exception of T1; Online Resource1 Table s8 c).

Blank-to-core ratios were analysed across assemblages [@fig-ratios b, Online Resource 1 Table s7]. In all cases, local cherts generally exhibit lower blank-to-core ratios than non-local cherts, though variability in core sample sizes may influence these patterns. For the Gravettian assemblages, the results reveal that non-local cherts show ratios above 7 (with the exception of T10), while local cherts have ratios located below 4. The Solutrean assemblage also shows higher ratios for non-local cherts (\>12), with local cherts showing ratios below 7. The Proto-Solutrean assemblage has most blank-to-core ratios below 10, except for non-local T7. However, non-local cores are poorly represented within the assemblage [@fig-ratios b, Online Resource 1 Table s3].

@@

The sizes of blanks and cores were explored through scatterplots of width and length [@fig-gravettian-plots; @fig-proto-plots; @fig-solutrean-plots a] and boxplots for weight [@fig-gravettian-plots; @fig-proto-plots; @fig-solutrean-plots b] and through non-parametric statistical tests; complete tables with descriptive statistics and non-parametric statistical test results can be found in Online Resource 1 Tables s9-11 and s12, respectively.

The scatterplots and boxplots show different patterns in the several analysed classes. In the Gravettian assemblage [@fig-gravettian-plots], non-local chert blanks are significantly smaller than local chert blanks in both length (\<25 mm) and weight (\<5 g), with p-values of 0.00009 and 0.012, respectively. While local chert blanks also cluster below 25 mm in width and length, they display greater variability and a higher number of outliers. The same patterns can be seen in the Proto-Solutrean assemblage [@fig-proto-plots], where length and weight also show statistically significant differences between chert types (0.046 and 0.004), with non-local types showing smaller values. In comparison, in the Solutrean assemblage [@fig-solutrean-plots], only non-local chert blank length seems to show statistically significant smaller values (p value of 0.0005) than local chert blanks, being mostly concentrated below \~30 mm, while local chert blank lengths are more variable. While width seems to be more variable in both local and non-local chert blanks, the statistical analysis shows that non-local chert width is significantly higher than local chert blanks (p value of 0.02).

@@

For cores, size and weight differences between local and non-local cherts are less pronounced than those observed in blanks. In the Gravettian assemblage [@fig-gravettian-plots], there is a cluster of non-local chert cores below the 25 mm width and length, while in general, local chert cores seem to be more variable. Especially regarding weight, T6 cores (non-local) show lighter cores (\~12 gr means) compared to local cherts. In the Solutrean assemblage [@fig-solutrean-plots], weight is the only measurement where differences can be seen between local and non-local chert cores; similarly to the Gravettian assemblage, T6 cores show slightly smaller cores (\~18 gr means) than local cherts (\~22 gr means). Despite these differences, the statistical tests applied to the dimensions between local and non-local chert cores show no statistically significant differences between the groups (p \> 0.05). The Proto-Solutrean assemblage [@fig-proto-plots] contains a limited sample of non-local chert cores (n=4), preventing clear pattern identification. However, as in other assemblages, local chert cores display a broad size range (\~15-50 mm in width/length, \~3-85 g in weight).

@@

Our results indicate that most non-local cherts were introduced to Vale Boi as cores or nodules, subsequently knapped on-site. However, an exception is seen in TL cherts (TL2-6, TL9-10, TL14; Online Resource 1 Table S13), which lack associated cores and knapping debris. This absence suggests that TL cherts were introduced to the site primarily as finished tools or blanks, aligning with long-distance individual provisioning models [@kuhn2004].

@@

# Statements and Declarations

## Funding

@@

## Competing Interests

>>>>

---
title: "From Stone to Tool: How Raw Materials Influenced Upper Paleolithic Technology in Southwestern Iberia (Vale Boi)"
author:
  - Joana Belmiro:
      correspondence: "yes"
      email: jfbelmiro@ualg.pt
      orcid: https://orcid.org/0000-0003-4951-2907
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
  - Nuno Bicho:
      orcid: https://orcid.org/0000-0001-9655-0549
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
  - Xavier Terradas:
      orcid: https://orcid.org/0000-0002-8000-5607
      institute:
        - Institución Milá y Fontanals de Investigación en Humanidades, Consejo Superior de Investigaciones Científicas, Barcelona, Spain
  - João Cascalheira:
      orcid: https://orcid.org/0000-0003-0321-8892
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
title-block-published: "Last updated"
format:
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
number-sections: true
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/archaeological-and-anthropological-sciences.csl" # Insert path for the bib-style
abstract: |
  The Upper Palaeolithic (UP) of westernmost Europe was marked by technological and cultural transformations and abrupt climatic shifts. The Iberian Peninsula, particularly southwestern Iberia, served as a refugia and key eco-cultural niche, making it a key region for UP studies. Vale Boi, the only site in southwestern Iberia with a nearly complete UP sequence, provides critical insights into technological, economical and mobility patterns over time. This study examines lithic raw material procurement and technological organisation at Vale Boi, focusing on the differences between local and non-local cherts. We analysed chert assemblages from Gravettian, Proto-Solutrean, and Solutrean levels, integrating previously published techno-typological and raw material macroscopic and petrographic data. Our results indicate that non-local cherts were intensively reduced. The presence of varied techno-typological classes suggests that non-local cherts were not exclusively transported as finished tools, instead highlighting their role as versatile, transportable raw material volumes. Gravettian occupations exhibited greater reliance on non-local cherts, suggesting short-term occupations and higher mobility, while Proto-Solutrean and Solutrean occupations showed increased dependence on local materials, suggesting long-term occupations. These findings expand our understanding of UP technological adaptations in southwestern Iberia, emphasizing the complex role of lithic resources in mobility, settlement, and social networks.

@@

In westernmost Europe, five main technocomplexes have been identified within the UP - Aurignacian, Gravettian, Proto-Solutrean, Solutrean, and Magdalenian - each contributing new evidence on resilient behaviours and cultural continuity. These technocomplexes are distinguished by regional adaptations as well as technological and subsistence innovations [@baker2024; @bicho2017; @bicho2011; @bicho2008; @cascalheira2017; @straus1991; @straus1993; @zilhão1997].

These innovations are believed to have resulted from human adaptations and resilience to the shifting conditions of the Late Pleistocene, which was marked by abrupt climatic events such as the Heinrich events and the Last Glacial Maximum (LGM) [@sanchezgoñi2010]. These climatic shifts had a profound impact on flora and fauna, with exception of spatially limited refugia, likely disrupting settlement patterns and the organisation of hunter-gatherer groups, ultimately leading to a process of cultural reorganisation [@bradtmöller2012; @morin2008; @schmidt2012], which has frequently been a focus of UP research.

@@

In this context, comparisons between Vale Boi and the Gravettian occupations in central Portugal and southern Spain revealed marked regional differences, suggesting more limited movement of people and information in southwestern Iberia. These differences were suggested to be related to Vale Boi’s peripheral location and the distinct ecological characteristics of westernmost Algarve [@marreiros2013]. In contrast, Proto-Solutrean evidence from the site indicates technological similarities and strong connections with central Portugal [@belmiro2021], while the Solutrean period stands out as a highly dynamic phase at Vale Boi, during which the site functioned as a key connection point between central Portugal and southern Spain, facilitating social networks and cultural exchanges [@bicho2017a; @cascalheira2013c].

@@

This paper presents a combined analysis of lithic technology and raw material sourcing, aiming to refine our understanding of technological and social adaptations throughout the UP in southwestern Iberia. By integrating archaeological and ethnographic perspectives, we test expectations regarding the organisation of hunter-gatherer technological systems and their implications for mobility, resource management, and social interactions. In doing so, we contribute to ongoing discussions about eco-cultural niches and the role of southwestern Iberia within broader UP social networks.

@@

Lithic technological organisation has long been a central focus in archaeology, offering critical insights into past human behaviour and adaptive strategies. It encompasses all stages of stone tool production and use, from raw material acquisition to discard. Nelson [-@nelson1991] defines it as the selection and integration of strategies for tool production, use, transport, and maintenance, while Kelly [-@kelly1988] emphasizes its spatial and temporal dimensions within cultural systems. These perspectives highlight the interrelationship between raw material sourcing, mobility, and settlement organisation—key factors in understanding hunter-gatherer adaptations.

@@

Vale Boi is located on the western coast of the Algarve region, southern Portugal, at a latitude of N 37º5'24.1" and longitude of W 8º48'32.4" [@gibajaFUNCIONINSTRUMENTOSLITICOS2006, the original site coordinates were converted from Degree/Minute to Degree/Minute/Second], within a small valley extending \~2 km southward to the Atlantic Ocean ([@fig-map] a). The site is nestled among limestone outcrops that create west- and southwest-facing rock shelters ([@fig-map] b). Covering more than 10,000 square meters along the valley slope, Vale Boi includes three main excavated areas, Slope, Terrace, and Shelter ([@fig-map] c), which were excavated between 2000 and 2019 [@bicho2007; @bicho2012; @bicho2013; @cascalheira2017; @cascalheira2013; @manne2012]. Based on the available raw material and technological evidence, particularly the macroscopic and petrographic data [@belmiro2025], the Terrace and Shelter areas were selected for the current study.

@@

The site’s lithic assemblages represent all stages of stone tool production, with raw material use varying by occupation. Technological analyses of these assemblages have provided insights into the adaptive strategies of UP hunter-gatherers at Vale Boi, revealing technological shifts across different cultural phases. Given the chosen sample for this study (see Section Datasets and analysis), we summarise here the results for the technological analysis of the Solutrean assemblages (levels A-C) of the Shelter [@cascalheira2010; @cascalheira2013a] and the Proto-Solutrean assemblages (levels 4E-5) of the Terrace [@belmiro2020; @belmiro2021]. The Gravettian occupations (levels 6-8) of the Terrace have been recently studied, but the results remain unpublished.

@@

Previous studies showed that the main raw materials used at the site are chert, quartz, greywacke and smaller amounts of chalcedony, with the percentages of these changing across occupations. Notably, the lower Proto-Solutrean levels show high frequencies of quartz [@belmiro2021], whereas chert becomes the dominant raw material in the Solutrean occupations of the Shelter [@cascalheira2010]. Likewise, different raw materials appear to have been used for specific purposes, with chert being particularly associated with the production of high percentage of blanks and retouched tools, and greywacke being extensively used for anvils [@horta2019].

Quartz and greywacke are locally available from streambeds near the site, whereas regional chert and chalcedony sources lie further afield, within a \~20 km radius [@pereira2016]. Our previous studies showed, as expected, the existence of local cherts in the assemblages ([@fig-cherts] a), originating from the Lower Jurassic (cherts) and Triassic (chalcedony) outcrops in proximity to the site [@belmiro2025]. Despite petrographic differences between cherts and chalcedony, for this study any chert varieties (e.g., chalcedony or jasper) are referred to as chert [@luedtke1978]. A significant portion of non-local cherts was also identified (@fig-cherts b-c), with some identified sources, such as the Cretaceous outcrops of central Portugal (Lisbon and Rio Maior areas) and the Upper Jurassic oolitic/peloidal cherts of the Betic Systems in southern Spain. This indicates varying degrees of contact or long-range mobility throughout the UP, likely influenced by shifts in settlement patterns and the duration of site occupation.

@@

Cascalheira et al. (2017) applied the Whole Assemblage Behavioural Index (WABI) [@riel-salvatore2004; @riel-salvatore2010] to reconstruct and compare mobility strategies at Vale Boi. By analysing lithic volumetric density and retouched tool frequencies from the Shelter’s Solutrean [@cascalheira2010; @cascalheira2013a] and Magdalenian [@mendonça2009] levels, as well as the Terrace’s Gravettian levels [@marreiros2015], the authors confirmed that most occupations corresponded to long-term residential base camps, composed of high-density assemblages with low degree of retouch, except for the Early Gravettian and Magdalenian occupations. Regarding the Terrace area, however, the results were based on previous excavations (\<2012), and the identification of new layers and reorganisation of the occupations is yet to be tested through the use of lithic techno-typological methods.

@@

One challenge of integrating datasets collected by different researchers is variability in methodologies and recorded variables. However, this study minimizes such inconsistencies by ensuring methodological alignment across datasets, as all data collection followed standardised attribute-based lithic analysis protocols. The lithic analysis datasets used in this study adopted an attribute-based methodology, which aims to describe morphological and metrical attributes of technological classes. The attributes analysed followed those present in specialised literature, such as Brézillon [-@brézillon1968], and [@tixier1980], paired with other Palaeolithic lithic attribute analyses [e.g., @bicho1992; @zilhão1997; @cascalheira2019; @tostevin2012].

After data collection, the databases were imported into R Programming Environment, where the information was processed through the creation of descriptive statistical analysis and writing of this article. Some differences in the collection process, especially between the study by Cascalheira [-@cascalheira2010] and the following studies, required homogenisation. For example, where in Cascalheira (2010) retouch was an independent variable (e.g., Class as flake, Retouch as Yes), in Belmiro et al. [-@belmiro2021], retouch was identified as Class variable. In this situation, the Solutrean dataset was modified, so that all artifacts with retouch were identified in the Class variable, instead of the original classification. All changes applied to the original data made for homogenisation between datasets can be consulted in the R Studio scripts in our online repository (<http://www.doi.org/10.17605/OSF.IO/UWMPK>).

To integrate technological and raw material data, we merged multiple datasets using the artifact identifier (“ID”), which corresponds to total station plot records. This identifier links artifacts to either precise excavation coordinates (for individually plotted pieces) or generalised spatial data (for sieved assemblages).

Given that the raw material analysis focused exclusively on individually piece-plotted artifacts (dimensions \>2 cm), the final dataset does not include all excavated material but retains the most diagnostically relevant pieces for analysis. As a result, the final dataset used in this study does not directly correspond to the full datasets analysed in the original studies ([@tbl-samples]). Similarly, the raw material study used only currently accessible artifacts from the assemblages. This is especially relevant for the Shelter Solutrean chert assemblage, where shatter is absent from the sample due to the lack of availability for study.

@@

Following data homogenization and integration, we conducted descriptive statistical analyses and non-parametric statistical analyses to examine correlations between raw material use and technological patterns and test expectations related to the technological organisation of local and non-local cherts (see [@tbl-expectations]), and between different UP occupations. To test these expectations, we evaluated 9 parameters for each dataset and assemblage; the parameters and their calculation/measurement methods are detailed in @tbl-parameters.

@@

The entirety of the R code used for the analysis, datasets, and visual representations contained in this paper can be accessed through our online repository (<http://www.doi.org/10.17605/OSF.IO/UWMPK>). We used the rrtools package by Marwick et al. [-@marwick2018] to create a research compendium and write a reproducible journal article. The provided files include the complete set of raw data used in the analysis, along with a custom R project [@wickham2015] containing the code required to generate all tables and figures. To enable maximum reuse, the code is made available under the MIT license, data under CC‐0, and figures under CC‐BY [additional details can be found in @Marwick2017].

@@

Results are structured according to the predefined analytical parameters, comparing local and non-local cherts ([@fig-cherts]) across different assemblages. A comprehensive description of each assemblage and detailed tables for all analyses are provided in Online Resource 1.

Blanks and blank fragments represent the most abundant lithic class in all assemblages, accounting for over 40% of artifacts across all chert types ([@fig-general-graphs]). As expected, cores are present in lower frequencies (\~3-8%) but are more numerous in local cherts than in non-local types and absent from the TL group. In the Proto-Solutrean assemblage, non-local chert cores are rare, with only T7-9 yielding a small number (n=4 complete cores, n=2 core fragments). Albeit in very small numbers in all assemblages (n\<10), core preparation and maintenance products are also present in local and non-local cherts, especially in the non-local types which show the larger sample number such as T6-7. The presence of knapping by-products such as shatter is also seen in both local and non-local cherts in the Gravettian (@fig-general-graphs a) and Proto-Solutrean (@fig-general-graphs b) assemblages. Retouched tools are present in all chert types across all assemblages. In general, local cherts exhibit lower retouch frequencies than non-local cherts, though the total number of retouched tools varies significantly between assemblages.

@@

The Gravettian assemblage exhibits the highest proportion of retouched tools (\~21%; [@fig-general-graphs] a), compared to \<10% in other assemblages. Tool diversity is greater among non-local cherts, which contain a high frequency of generalized tools (e.g., notches, retouched pieces) alongside hunting implements such as backed bladelets. The abundance of burins and burin spalls suggests they may have functioned as cores, indicating a technological strategy rather than solely a focus on retouch (Online Resource 1 Tables s1 and s3). In comparison to the Gravettian, other assemblages show smaller numbers of retouched tools. While in the Proto-Solutrean (Online Resource 1 Table s5) no significant difference was identified between local and non-local retouched tools, in the Solutrean assemblage (Online Resource 1 Table s7) both tool diversity and tool specialization are higher in local cherts.

Retouched tools were further explored through the tools-to-debitage ratio (@fig-ratios a, Online Resource 1 Table s8). In general, the results show that non-local chert types have higher tool-to-debitage ratios than local types, albeit with differences between assemblages. The Gravettian assemblage shows the highest tool-to-debitage ratios compared to the remaining assemblages; non-local cherts show ratios above 0.9, while local cherts show values below 0.5. In the Proto-Solutrean non-local cherts show ratios between 0.3 to 0.5, while local cherts show ratios below 0.3. This pattern is less clear in the Solutrean assemblage since the ratios are the lowest of all assemblages. Non-local cherts show ratios above 0.1 (with exception of T8) while local cherts show ratios below 0.1 (with exception of T1; Online Resource1 Table s8 c).

@@

Blank-to-core ratios were analysed across assemblages (@fig-ratios b, Online Resource 1 Table s9). In all cases, local cherts generally exhibit lower blank-to-core ratios than non-local cherts, though variability in core sample sizes may influence these patterns. For the Gravettian assemblages, the results reveal that non-local cherts show ratios above 7 (with the exception of T10), while local cherts have ratios located below 4. The Solutrean assemblage also shows higher ratios for non-local cherts (\>12), with local cherts showing ratios below 7. The Proto-Solutrean assemblage has most blank-to-core ratios below 10, except for non-local T7. However, non-local cores are poorly represented within the assemblage (@fig-ratios b, Online Resource 1 Table s4).

The sizes of blanks and cores were explored through scatterplots of width and length (@fig-gravettian-plots, @fig-proto-plots, @fig-solutrean-plots a) and boxplots for weight (@fig-gravettian-plots, @fig-proto-plots, @fig-solutrean-plots b) and through non-parametric statistical tests; complete tables with descriptive statistics and non-parametric statistical test results can be found in Online Resource 1 Tables s10-12 and s13, respectively.

The scatterplots and boxplots show different patterns in the several analysed classes. In the Gravettian assemblage ([@fig-gravettian-plots]), non-local chert blanks are significantly smaller than local chert blanks in both length (\<25 mm) and weight (\<5 g), with p values of 0.00009 and 0.012, respectively. While local chert blanks also cluster below 25 mm in width and length, they display greater variability and a higher number of outliers. The same patterns can be seen in the Proto-Solutrean assemblage ([@fig-proto-plots]), where length and weight also show statistically significant differences between chert types (0.046 and 0.004), with non-local types showing smaller values. In comparison, in the Solutrean assemblage ([@fig-solutrean-plots]), only non-local chert blank length seems to show statistically significant smaller values (p value of 0.0005) than local chert blanks, being mostly concentrated below \~30 mm, while local chert blank lengths are more variable. While width seems to be more variable in both local and non-local chert blanks, the statistical analysis shows that non-local chert width is significantly higher than local chert blanks (p value of 0.02).

@@

For cores, size and weight differences between local and non-local cherts are less pronounced than those observed in blanks. In the Gravettian assemblage ([@fig-gravettian-plots]), there is a cluster of non-local chert cores below the 25 mm width and length, while in general, local chert cores seem to be more variable. Especially regarding weight, T6 cores (non-local) show lighter cores (\~12 gr means) compared to local cherts. In the Solutrean assemblage ([@fig-solutrean-plots]), weight is the only measurement where differences can be seen between local and non-local chert cores; similarly to the Gravettian assemblage, T6 cores show slightly smaller cores (\~18 gr means) than local cherts (\~22 gr means). Despite these differences, the statistical tests applied to the dimensions between local and non-local chert cores show no statistically significant differences between the groups (p \> 0.05). The Proto-Solutrean assemblage ([@fig-proto-plots]) contains a limited sample of non-local chert cores (n=4), preventing clear pattern identification. However, as in other assemblages, local chert cores display a broad size range (\~15-50 mm in width/length, \~3-85 g in weight).

@@

Our results indicate that most non-local cherts were introduced to Vale Boi as cores or nodules, subsequently knapped on-site. However, an exception is seen in TL cherts (TL2-6, TL9-10, TL14; Online Resource 1 Table S2), which lack associated cores and knapping debris. This absence suggests that TL cherts were introduced to the site primarily as finished tools or blanks, aligning with long-distance individual provisioning models [@kuhn2004].

@@

# Statements and Declarations {.unnumbered}

## Funding {.unnumbered}

@@

## Competing Interests {.unnumbered}
